Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents

### 2025-07-24 doc/cicm25-paper/paper.tex
some edits

  Semantic annotations enable services that make the knowledge in documents more
  accessible.  In the case of \ALeA, a learning assistant we use as a case study, they
  enable, for example, guided tours and practice problems tailored to the learner's
  competencies.  However, the cost of semantic authoring is substantial and tool support
  is limited.

  In this paper we discuss the nature of the semantic authoring problem and argue that,
  while sharing aspects of classical (informal) authoring and formal authoring, it is a
  problem of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
    semantic references to a flexiformal content commons \edited{and increases} annotation
  productivity by almost an order of magnitude.
\end{abstract}
% \setcounter{tocdepth}{3}
% \tableofcontents

\section{Introduction}
Arguably, handling large document collections is a key factor in our knowledge-driven society and economy.
There are two main contenders for machine support in this area:
symbolic/logic-based technologies and \edited{statistical}/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, \edited{offer precision and explainability%
%out of the box%
, but face scalability challenges}
-- in particular because of the cost of semantic authoring, which we discuss in this paper.

As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
precision and explainability required by the ethics of teaching --
while the scope of specific domains is just narrow enough that symbolic approaches are viable.
Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
depends on semantic annotations.
These can be embedded in \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
In this paper, we will mostly focus on annotating terms with the semantic concepts they refer to,
but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
or annotating the defining phrase (definiens) in concept definitions,
and could, in principle, even be used for full formalization.
As \sTeX supports the spectrum from informal (un-annotated) to fully formalized content,
we say that \sTeX content is \textbf{flexiformal}.

@@

\paragraph{Acknowledgments}
The work reported in this article was conducted as part of the VoLL-KI project (see
\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMBF) under
grant 16DHBKI089.

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from the \sTeX/\ALeA world.
    % We will use it to sharpen our intuitions of the challenges involved and to ground the solution we propose.

\subsection{Running Example: Annotating ``terminal''}
With \sTeX, authors can add semantic references to technical terms,
which enables straightforward learning support services like
showing the definition of a term, but also informs the system
about the content/prerequisites of a learning object.
Adding these references is a key practical problem when converting presentational {\LaTeX}
course materials into semantic documents.

@@

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both). Concretely this is about converting the unannotated {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
\edited{
where the author has annotated the word ``terminal'' with the concept it refers
to via} the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
In \sTeX, \edited{concepts} are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a \textbf{symbol name}.
 % Every symbol can have one or more verbalizations. %: words of the domain jargon used to refer to the concept or
\edited{
A \textbf{verbalization} is the natural language rendering of a concept.
% object.
Verbalizations are, of course, language-dependent and can vary between subdomains and communities of
practice.
}
In our example, the verbalization ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
% in \Cref{lst:losa}).

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF

>>>>

  Semantic annotations enable services that make the knowledge in documents more
  accessible.  In the case of \ALeA, an adaptive learning assistant we use as a case study, they
  enable, for example, guided tours and practice problems tailored to the learner's
  competencies.  However, the cost of semantic authoring is substantial and tool support
  is limited.

  In this paper we discuss the nature of the semantic authoring problem and argue that,
  while sharing aspects of both classical (informal) authoring and formal authoring, it is
  a problem of its own that needs special editing support facilities.  We present and
  evaluate a simple utility that supports the process of annotating technical terms with
  semantic references to a flexiformal content commons and increases annotation
  productivity by almost an order of magnitude.
\end{abstract}
% \setcounter{tocdepth}{3}
% \tableofcontents

\section{Introduction}
Arguably, handling large document collections is a key factor in our knowledge-driven
society and economy.  There are two main contenders for machine support in this area:
symbolic/logic-based technologies and statistical/machine-learning-based AI, e.g.\ via
LLMs or chatbots.  They have complementary strengths and challenges: ML-based approaches
can be trained on all data of the Internet, but face challenges in precision and
explainability.  Symbolic technologies, on the other hand, offer precision and
explainability, but face scalability challenges -- in particular because of the cost of
semantic authoring, which we discuss in this paper.

As a case study, we will use adaptive learning assistants for STEM subjects in higher
education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.  We contend that
machine-learning based technologies are largely unsuitable for this task -- at least until
they reach the precision and explainability required by the ethics of
teaching\edited{\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
    classifies education as a high-risk area and thus mandates high requirements of
    explainability, accountability, and human oversight for educational AI applications.}}
-- while the scope of specific domains is just narrow enough that symbolic approaches are
viable.  Conventional (\LaTeX-based) lecture materials can essentially be directly
imported into \ALeA, but its ability to generate learner-adaptive learning objects --
instrumented with learning support interactions -- depends on semantic annotations.  These
can be embedded in \LaTeX\ documents using the \sTeX
package~\cite{MueKo:sdstex22,sTeX:github:on}.  In this paper, we will mostly focus on
annotating terms with the semantic concepts they refer to, but \sTeX supports a much wider
range of annotations, such as annotating problems with the learning \edited{prerequisites
  and} objectives or annotating the defining phrase (definiens) in concept definitions,
and could, in principle, even be used for full formalization.  As \sTeX supports the
spectrum from informal (un-annotated) to fully formalized content, we say that \sTeX
content is \textbf{flexiformal}.

@@

\paragraph{Acknowledgments}
The work reported in this article was conducted as part of the VoLL-KI project (see
\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMFTR) under
grant 16DHBKI089.

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from
the \sTeX/\ALeA world.
% We will use it to sharpen our intuitions of the challenges involved and to ground the
% solution we propose.

@@

\subsection{Running Example: Annotating ``terminal''}
With \sTeX, authors can add semantic references to technical terms,
which enables straightforward learning support services like
showing the definition of a term, but also informs the system
about the content/prerequisites of a learning object.
Adding these references is a key practical problem when converting presentational {\LaTeX}
course materials into semantic documents.
\begin{figure}[h]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

Take for instance a learning object like the paragraph in \Cref{fig:lo}\ednote{MK: in
  diesem Bild sollten keine blauen/annotierten Wörter sein}, where we want to annotate the
term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The
latter can either be defined earlier in the course or in the domain model (or
both). Concretely this is about converting the unannotated {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated 
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
]
\importmodule[smglom/search]{mod?search-problem}
[...]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
where the author has annotated the word ``terminal'' with the concept it refers to via the
\sTeX macro
\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
In \sTeX, concepts are represented by \textbf{symbols}, which are identified globally and
unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a
\textbf{symbol name}.  A \textbf{verbalization} is the natural language rendering of a
concept.  Verbalizations are, of course, language-dependent and can vary between
subdomains and communities of practice, even if the underlying symbol is the same.  In our
example, the verbalization ``terminal [state]'' was annotated with the symbol with name
\lstinline|goal state|. It comes from the module with name \lstinline|search-problem| in
the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices to use a relative URI in the \lstinline|\sr| annotation. If the symbol name
and the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
% in \Cref{lst:losa}).

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF diesem Bild
% LocalWords:  sollten keine blauen annotierten Wörter sein
